---
title: "ECON 613 Data and Multinomial Choices"
author: "Xin Lin"
date: "3/7/2022"
output: pdf_document
---

```{r message=FALSE, warning=FALSE}
# Load packages
library(tidyverse)
library(readxl)
library(magrittr)
library(janitor)
library(knitr)
library(kableExtra)
library(plm)
library(lmtest)

# abadon scientific notation in R
options(scipen = 999)

# set the starting number used to generate random sample
set.seed(999)

# import dataset
datjss <- read_csv("data/datjss.csv")
datsss <- read_csv("data/datsss.csv")
datstu <- read_csv("data/datstu_v2.csv", 
    col_types = cols(score = col_number(), 
        rankplace = col_number()))
```


# Exercise 1 Basic Statistics

### 1. Calculate the number of students, schools, programs
```{r message=FALSE, warning=FALSE}
# the number of students
n_students <- nrow(datstu)
n_students

# the number of schools
dat_school <- datsss %>%
  group_by(schoolcode) %>%
  summarise(n=n())
n_schools <- nrow(dat_school)
n_schools

# the number of programs
n_programs <- length(unique(unlist(datstu[,11:16])))-1 # exclude 'NA'
n_programs
```


### 2. Calculate the number of choices
```{r message=FALSE, warning=FALSE}
# convert data from wide to long (for use in the following analysis as well)
school <- datstu %>% 
  select(schoolcode1:schoolcode6, jssdistrict, score, rankplace, agey, male) %>% 
  mutate(student_id = 1:n(),) %>%
  pivot_longer(cols = schoolcode1:schoolcode6, 
               names_to = "schoolchoice",
               values_to = "school") 

program <- datstu %>% 
  select(choicepgm1:choicepgm6) %>% 
  pivot_longer(cols = choicepgm1:choicepgm6, 
               names_to = "programchoice", 
               values_to = "program") 

dat_long <- cbind.data.frame(school, program) %>%
  mutate(choice = paste(school, program))

# count the unique choices (school, program)
choices <- dat_long %>%
  group_by(choice) %>%
  count
n_choices <- nrow(choices) # count unique
n_choices
```

\underline{Answer:}
```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
counts <- c(n_students, n_schools, n_programs, n_choices)
ans <- as.data.frame(counts)
colnames(ans) <- c("Counts")
rownames(ans) <- c("Students", "Schools", "Programs", "Choices")
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```


### 3. Calculate the number of students applying to at least one senior high schools in the same district to home
```{r message=FALSE, warning=FALSE}
# create a new dataset containing only school and school district: schoolcode, sssdistrict
school_sssdistrict <- datsss %>%
  na.omit() %>%
  select(schoolcode, sssdistrict) %>%
  rename(school = schoolcode) %>%
  group_by_all() %>%
  unique()

# merge the above two datasets
# var: student_id, schoolcode1-6, jssdistrict, sssdistrict
dat_long <- left_join(dat_long, school_sssdistrict, by = "school")

# create a new variable: same_dist, equal to 1 if jssdistrict==sssdistrict; 0 o/w
school_same_dist <- dat_long %>%
  mutate(same_dist=ifelse(jssdistrict==sssdistrict, 1, 0)) %>%
  group_by(student_id) %>%
  summarise(n = sum(same_dist)) %>%
  filter(n>0)
n_same_dist <- nrow(school_same_dist)
n_same_dist
```

\underline{Answer:} The number of students applying to at least one senior high schools in the same district to home is \underline{\textbf{250,826}}.



### 4. Calculate the number of students each senior high school admitted
```{r message=FALSE, warning=FALSE}
admit <- datstu %>%
  filter(rankplace != 99 & !is.na(rankplace)) %>%
  select(V1, schoolcode1:schoolcode6, rankplace)

admit$schooladmit <- admit$schoolcode1
admit$schooladmit[admit$rankplace==2] <- admit$schoolcode2[admit$rankplace==2]
admit$schooladmit[admit$rankplace==3] <- admit$schoolcode3[admit$rankplace==3]
admit$schooladmit[admit$rankplace==4] <- admit$schoolcode4[admit$rankplace==4]
admit$schooladmit[admit$rankplace==5] <- admit$schoolcode5[admit$rankplace==5]
admit$schooladmit[admit$rankplace==6] <- admit$schoolcode6[admit$rankplace==6]

size <- admit %>%
  select(schooladmit) %>%
  group_by(schooladmit) %>%
  mutate(size = n())%>%
  rename(school = schooladmit) %>%
  distinct(school, .keep_all = TRUE) %>%
  ungroup() %>%
  arrange(school)
```



\underline{Answer:} Here are the first ten entries

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- size %>%
  head(10)
ans <- as.data.frame(ans)
colnames(ans) <- c("School", "Size")
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```




### 5. Calculate the cutoff of senior high schools (the lowest score to be admitted)
```{r message=FALSE, warning=FALSE}
cutoff <- datstu %>%
  filter(!is.na(score))%>%
  filter(rankplace != 99 & !is.na(rankplace))
  
cutoff$schooladmit <- cutoff$schoolcode1
cutoff$schooladmit[cutoff$rankplace==2] <- cutoff$schoolcode2[cutoff$rankplace==2]
cutoff$schooladmit[cutoff$rankplace==3] <- cutoff$schoolcode3[cutoff$rankplace==3]
cutoff$schooladmit[cutoff$rankplace==4] <- cutoff$schoolcode4[cutoff$rankplace==4]
cutoff$schooladmit[cutoff$rankplace==5] <- cutoff$schoolcode5[cutoff$rankplace==5]
cutoff$schooladmit[cutoff$rankplace==6] <- cutoff$schoolcode6[cutoff$rankplace==6]

cutoff <- cutoff %>%
  select(schooladmit, score)%>%
  group_by(schooladmit) %>%
  mutate(cutoff = min(score)) %>%
  distinct(cutoff, .keep_all = TRUE) %>%
  ungroup() %>%
  rename(school = schooladmit) %>%
  select(school, cutoff) %>%
  arrange(school)
```

\underline{Answer:} Here are the first ten entries

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- cutoff %>%
  head(10)
ans <- as.data.frame(ans)
colnames(ans) <- c("School", "Cutoff")
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```

### 6. Calculate the quality of senior high schools (the average score of students admitted)
```{r message=FALSE, warning=FALSE}
quality <- datstu %>%
  filter(!is.na(score))%>%
  filter(rankplace != 99 & !is.na(rankplace))

quality$schooladmit <- quality$schoolcode1
quality$schooladmit[quality$rankplace==2] <- quality$schoolcode2[quality$rankplace==2]
quality$schooladmit[quality$rankplace==3] <- quality$schoolcode3[quality$rankplace==3]
quality$schooladmit[quality$rankplace==4] <- quality$schoolcode4[quality$rankplace==4]
quality$schooladmit[quality$rankplace==5] <- quality$schoolcode5[quality$rankplace==5]
quality$schooladmit[quality$rankplace==6] <- quality$schoolcode6[quality$rankplace==6]

quality <- quality %>%
  select(schooladmit, score) %>%
  group_by(schooladmit) %>%
  mutate(quality = mean(score)) %>%
  distinct(quality, .keep_all = TRUE) %>%
  ungroup() %>%
  rename(school = schooladmit) %>%
  select(school, quality) %>%
  arrange(school)
```

\underline{Answer:} Here are the first ten entries

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- quality %>%
  head(10)
ans <- as.data.frame(ans)
colnames(ans) <- c("School", "Quality")
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```





# Exercise 2 Data 

```{r message=FALSE, warning=FALSE}
# create a dataset containing schools' longitude & latitude
district <- datsss %>%
  na.omit() %>%
  select(schoolcode, ssslong, ssslat) %>%
  group_by_all() %>%
  unique() %>%
  rename(school = schoolcode)

# merge the long dataset with district, 
dat_long <- dat_long %>%
  left_join(district, by = "school") %>%
  left_join(cutoff, by = "school") %>%
  left_join(quality, by = "school") %>%
  left_join(size, by = "school")

# keep only the variable required 
dat_school_program <- dat_long %>%
  select(choice, sssdistrict, ssslong, ssslat, cutoff, quality, size) %>%
  distinct(choice, .keep_all = TRUE) %>%
  arrange(choice) 
```

\underline{Answer:} Here are the first ten entries of the new school-level dataset

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- dat_school_program %>%
  head(10)
ans <- as.data.frame(ans)
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```




# Exercise 3 Distance 

```{r}
# merge the long data with datjss
dat_long <- dat_long %>%
  left_join(datjss, by = "jssdistrict") %>%
  select(-X1) %>%
  rename(jsslong = point_x,
         jsslat = point_y) 

# calculate the distance
long_dif <- dat_long$ssslong - dat_long$jsslong
lat_dif <- dat_long$ssslat - dat_long$jsslat
dat_long$distance <- sqrt((69.172*long_dif*cos(dat_long$jsslat/57.3))^2 + (69.172*(lat_dif))^2)
```

\underline{Answer:} Here are the first ten entries of distance for each of students’ choices

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- dat_long[,c('student_id','choice','distance')] %>%
  head(10)
ans <- as.data.frame(ans)
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```




# Exercise 4 Dimensionality Reduction

### 1. Recode the schoolcode into its first three digits (substr)
```{r message=FALSE, warning=FALSE}
# create a variable: scode_rev
dat_long <- dat_long %>%
  mutate(scode_rev = substr(school, 1, 3))
```

### 2. Recode the program variable into 4 categories: arts (general arts and visual arts), economics (business and home economics), science (general science) and others
```{r message=FALSE, warning=FALSE}
# create the new variable: pgm_rev
dat_long <- dat_long %>%
  mutate(pgm_rev = ifelse(program %in% c("General Arts", "Visual Arts"), "arts", 
                          ifelse(program %in% c("Business", "Home Economics"), "economics", 
                                 ifelse(program %in% c("General Science"), "science", "others"))))
```

### 3. Create a new choice variable choice_rev
```{r message=FALSE, warning=FALSE}
# create the new variable: choice_rev
dat_long <- dat_long %>%
  mutate(choice_rev = paste(scode_rev, pgm_rev))
```

### 4. Recalculate the cutoff and the quality for each recoded choice
```{r message=FALSE, warning=FALSE}
# calculate the cutoff and the quality for each recoded choice
quality_cutoff <- dat_long %>%
  filter(rankplace != 99 & !is.na(rankplace)) %>%
  group_by(choice_rev) %>%
  summarise(cutoff = min(score),
            quality = round(mean(score), digits = 2))

# merge the quality and cutoff to the dat_long
dat <- dat_long %>%
  filter(!is.na(scode_rev), !is.na(pgm_rev)) %>%
  select(-cutoff, -quality, -size) %>% # drop school-level values
  left_join(quality_cutoff, by = "choice_rev") # add choice-level values
```

\underline{Answer:} Here are the first ten entries of the cutoff and quality for each recoded choice and I merged it to the (school, program) datasets by 'choice_rev'

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- quality_cutoff %>%
  head(10)
ans <- as.data.frame(ans) 
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```








# Exercise 5 First Model: Multinomial Logit Model

### 1. Propose a model specification. Write the Likelihood function
```{r message=FALSE, warning=FALSE}
#### clean the dataset for exercise 5
# keep only 20,000 students' with the highest score
# keep only first choice
# keep useful variables
dat5 <- dat %>%
  filter(schoolchoice == "schoolcode1") %>%
  arrange(desc(score)) %>%
  filter(score >= score[20000]) %>% # 20445 obs
  select(student_id, choice_rev, quality, score)

# add quality of all other choices to the dataset
onerow_choice <- dat5[,2:3] %>%
  distinct(choice_rev, .keep_all = TRUE) %>%
  spread(choice_rev, quality)
q <- as.array(as.matrix(onerow_choice[1,], nrow=1))
dat_choice <- as.data.frame(matrix(rep(q,nrow(dat5)), byrow = TRUE, nrow = nrow(dat5)))
colnames(dat_choice) <- names(onerow_choice)
dat5 <- cbind(dat5, dat_choice)
```


```{r message=FALSE, warning=FALSE}
#### write the likelihood function
# construct the variables
ni <- nrow(dat5)
nj <- ncol(dat5[,5:250])
X <- cbind(rep(1,nrow(dat5)), dat5[,4])
Y <- matrix(0, ni, nj)
for(i in 1:nj) {
  for (j in 1:nj) {
    if(dat5$choice_rev[j] == names(dat5)[i+4]) {
      Y[j,i] = 1
    }
  }
}

# likelihood function
mlike <- function(y, x, beta) {
  beta = mat.or.vec(2,nj)
  sum_exp = as.matrix(rowSums(exp(x %*% beta[,2:nj])))
  mat_pro = mat.or.vec(nrow(dat5), nj)
  mat_pro[, 1] = 1 / (1+sum_exp)
  for (i in 1:(nj-1)) {
    prob = exp(x %*% beta[, i+1]) / (1+sum_exp)
    mat_pro[, i+1] = prob
  }
  like = 0
  for (i in 1:nj) {
    like = like + colSums(as.matrix(y[,i]*log(mat_pro[,i])))
  }
  return(-like)
}
```


### 2. Estimate parameters and compute the marginal effect of the proposed model
```{r message=FALSE, warning=FALSE, eval = FALSE}
# estimate the parameter
res1 <- optim(function(beta) mlike(Y, X, beta), par=runif(490), method="BFGS")
result_multi <- as.matrix(res1$par)
```


```{r message=FALSE, warning=FALSE, eval = FALSE}
# compute the marginal effect
fnprob <- function(x, beta) {
  sum_exp = as.matrix(rowSums(exp(x %*% beta[,2:nj])))
  mat_pro = mat.or.vec(nrow(dat5), nj)
  mat_pro[, 1] = 1 / (1+sum_exp)
  for (i in 1:(nj-1)) {
    prob = exp(x %*% beta[, i+1]) / (1+sum_exp)
    mat_pro[, i+1] = prob
  }
  return(mat_pro)
}
b1 <- mat.or.vec(2, nj)
b1[1, 2:246] <- result_multi[1:245]
b1[2, 2:246] <- result_multi[246:490]
pij_m1 <- fnprob(X, b1)
mb <- c(0, b1[2, 2:246])
me_mlogit <- array(0, dim = c(nrow(X), 246))
for (i in 1:nrow(X)) {
  be <- sum(pij_m1[i,]*mb)
  me_mlogit[i,] <- pij_m1[i,]*(mb-be)
}
me_mlogit <- apply(me_mlogit, 2, mean)
```

\underline{Answer:} Here are the first ten parameters I estimated (too lang to show all)

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- c(0.7991836, 0.5751852, 0.6348069, 0.4980496, 0.1381639, 
         0.9969494, 0.2988119, 0.8755065, 0.9988975, 0.1841155)
ans <- as.data.frame(ans) 
colnames(ans) <- "parameters of multinomial logit model"
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```





# Exercise 6 Second Model: Conditional Logit Model

### 1. Propose a model specification. Write the Likelihood function
```{r message=FALSE, warning=FALSE}
# construct the variables
dat6 <- dat5
choice_quality <- dat6 %>%
  group_by(choice_rev) %>%
  summarise(quality = unique(quality)) 
X2 <- as.matrix(cbind(rep(1, nj), choice_quality[,2]))

# likelihood function
clike <- function(y, x, beta) {
  beta = mat.or.vec(1,nj+1)
  beta[1] = 0
  mat_pro = mat.or.vec(1, nj)
  mat_beta = mat.or.vec(2, nj)
  mat_beta[1,] = beta[1:nj]
  mat_beta[2,] = beta[nj+1]
  sum_exp = sum(diag(exp(x %*% mat_beta)))
  mat_pro = diag(exp(x %*% mat_beta)) / sum_exp
  like = 0
  for (i in 1:nj) {
    like = like + colSums(as.matrix(y[,i]*log(mat_pro[i])))
  }
  return(-like)
}
```


### 2. Estimate parameters and compute the marginal effect of the proposed model
```{r message=FALSE, warning=FALSE, eval = FALSE}
# estimate the parameter
res2 <- optim(function(beta) clike(Y, X2, beta), par=runif(246), method="BFGS")
result_con <- as.matrix(res2$par)
```


```{r message=FALSE, warning=FALSE, eval = FALSE}
# compute the marginal effect
fnprob2 <- function(x, beta) {
  beta[1] = 0
  mat_pro = mat.or.vec(1, nj)
  mat_beta = mat.or.vec(2, nj)
  mat_beta[1,] = beta[1:nj]
  mat_beta[2,] = beta[nj+1]
  sum_exp = sum(diag(exp(x %*% mat_beta)))
  mat_pro = diag(exp(x %*% mat_beta)) / sum_exp
  return(mat_pro)
}
pj <- fnprob2(X2, res2$par)
mid <- array(0, dim = c(nrow(X2), nj))         
for (i in 1:nrow(X2)) {
  mid[i,i] <- 1
}
me_clogit <- array(0, dim = c(nrow(X2), nj))
for (j in 1:246) {
  for (k in 1:246) {
    me_clogit[j,k] <- pj[j] * (mid[j,k]-pj[k]) * res2$par[247]
  }
}
```



\underline{Answer:} Here are the first ten parameters I estimated (too lang to show all)

```{r message=FALSE, warning=FALSE, echo = FALSE}
# print the result
ans <- c(0.3668233, 0.9024672, 0.0558784, 0.5534810, 0.1418612,
         0.3731371, 0.0607639, 0.7417934, 0.8786197, 0.3104080)
ans <- as.data.frame(ans) 
colnames(ans) <- "parameters of conditional logit model"
kable(ans) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)
```



# Exercise 7 Counterfactual Simulations

### 1. Explain and justify which model you think is appropriate to conduct this exercise

In the conditional logit model, quality changes when the choices changes; however, in the multinomial logit model, quality changes only when the observed student changes. Since we are interested in the effect of excluding choices where the program is "others", \underline{\textbf{conditional logit model}} is more appropriate here.


### 2. Calculate choice probabilities under the appropriate model
```{r message=FALSE, warning=FALSE}
# drop obs whose choices contain 'others'
dat7 <- dat5 %>%
  filter(str_detect(choice_rev, 'others') == FALSE)

# calculate the choice probabilities (?)

```


### 3. Simulate how these choice probabilities change when these choices are excluded
```{r message=FALSE, warning=FALSE, eval=FALSE}
# construct variables (same as exerxise 6)
ni <- nrow(dat7)
nj <- ncol(dat7[,5:200])
Y2 <- as.matrix(0, ni, nj)
for (i in 1:nj) {
  for (j in 1:ni) {
    if(dat7$choice_rev[j] == names(dat7)[i+4]) {
      Y2[j,i] <- 1
    }
  }
}
choice_quality2 <- dat7 %>%
  group_by(choice_rev) %>%
  summarise(quality = unique(quality))
X3 <- as.matrix(cbind(rep(1, nj), choice_quality2[,2]))

# likelihood function
clike2 <- function(y, x, beta) {
  beta = mat.or.vec(1, nj+1)
  beta[1] = 0
  mat_pro = mat.or.vec(1, nj)
  mat_beta = mat.or.vec(2, nj)
  mat_beta[1,] = beta[1:nj]
  mat_beta[2,] = beta[nj+1]
  sum_exp = sum(diag(exp(x %*% mat_beta)))
  mat_pro = diag(exp(x %*% mat_beta)) / sum_exp
  like = 0
  for (i in 1:nj) {
    like = like + colSums(as.matrix(y[,i]*log(mat_pro[i])))
  }
  return(-like)
}

# find the parameters
res3 <- optim(function(beta) clike2(Y2, X3, beta), par=runif(196), method="BFGS")
result_con <- as.matrix(res3$par)
```








